---
title: "Red-backed salamander  and distribution in north-central Massachusetts"
author: "Joshua A. Harkness"
date: "2024-02-13"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages(gridExtra)
library(tidyverse)
library(broom)
library(stringr)
library(sf)
library(ggspatial)
library(readxl)
library(GGally)
library(gridExtra)
library(tidymodels)
library(MASS)
library(simfit)
library(ggpmisc)
library(iNEXT)
library(pscl)
library(cowplot)
```

## Load Data
```{r load-datafiles}
plots <- read_csv("/cloud/project/data/hf131-01-plot.csv")
aco <- read_csv("/cloud/project/data/hf131-04-aco.csv")
pc <- read_csv("/cloud/project/data/hf131-05-pc-abund.csv")
herps <- read_csv("/cloud/project/data/hf131-06-herps.csv")
mass_towns <- st_read("/cloud/project/data/Community Boundaries (Towns) from Survey Points.shp")
coords <- read_csv("/cloud/project/data/coords.csv")
```
## Map plots
```{r transform-crs-mass_towns}
mass_towns <- st_transform(mass_towns, "+init=epsg:4326")
```

```{r plot-mass_towns-coords}
state_map <- ggplot(mass_towns) +
  geom_sf() +
  geom_point(data = coords, (aes(x = long_dd, y = lat_dd)), alpha = 0.5) +
  labs(x = "",
       y = "") +
   theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
   ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    pad_x = unit(0, "in"), pad_y = unit(0.2, "in"),
    style = ggspatial::north_arrow_fancy_orienteering(
      fill = c("grey40", "white"),
      line_col = "grey20",
      text_family = "ArcherPro Book")) +
  ggspatial::annotation_scale(
    location = "bl",
    bar_cols = c("grey60", "white"),
    text_family = "ArcherPro Book")

site_map <- ggplot(mass_towns) +
  geom_sf() +
  geom_point(data = coords, (aes(x = long_dd, y = lat_dd)), alpha = 0.5) +
  labs(x = "Long",
       y = "Lat",
       caption = "WGS 1984") +
  xlim(-72.4, -71.85) +
  ylim(42.4, 42.7) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
   ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    pad_x = unit(0, "in"), pad_y = unit(0.2, "in"),
    style = ggspatial::north_arrow_fancy_orienteering(
      fill = c("grey40", "white"),
      line_col = "grey20",
      text_family = "ArcherPro Book")) +
  ggspatial::annotation_scale(
    location = "bl",
    bar_cols = c("grey60", "white"),
    text_family = "ArcherPro Book") +
  annotate("text", x = c(-71.9, -72.12, -72.27, -72.12, -72.13, -72.09, -72.23, -72.36, -72.37, -72.38), y = c(42.52, 42.44, 42.49, 42.54, 42.62, 42.68, 42.69, 42.7, 42.66, 42.6), label = c("Mount Wachusett", "Slab City", "Tom Swamp", "Prospect Hill", "Millers River", "Beryl Hill", "Warwick SF", "Mount Grace SF", "Northfield SF", "Erving SF"))

state_map
site_map
```

## P. cinereus 
### Calculate  per cover board

```{r select-vars-aco}
aco <- aco %>%
  dplyr::select(plot, asn, aco, pca, osn, other.sp)
```

```{r calc-abund-aco}
aco_abund <- aco %>%
  group_by(plot, asn, aco) %>%
  summarize(n_ind = sum(pca), .groups = "keep")

aco_abund
```
Still getting 241 detections of P. cinereus, but when grouping by `plot` and `asn`, we get 310 cover board observations instead of the original 300 -- not sure why it's doing this, but presumbly it's only adding a ten zeros to the dataframe, so I'm going to let it go at that.

```{r join-aco_abund-plots}
df <- left_join(aco_abund, plots)
df

write_csv(df, "/cloud/project/data/df.csv")
```
```{r glimpse-df}
glimpse(df)
sum(df$n_ind)
```
Now 310 cover board observations, still with 241 P. cinereus detections.
```{r check-unique-n_ind-vals}
unique(df$n_ind)
```

```{r}
df %>%
  ggplot(aes(x = n_ind)) +
  geom_bar(fill = "seagreen4", color = "black") +
  labs(title = "P. cinereus abundance distribution",
       x = "Abundance per cover board",
       y = "Count")
```

#### Forest type comparisons


```{r plot-violin1}
violin1 <- df %>%
  ggplot(aes(x = ft, y = n_ind, fill = ft)) +
  geom_violin() +
  labs(title = "P. cinereus  by forest type",
       x = "Forest type",
       y = " per cover board") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("Mixed deciduous", "Hemlock")) +
  ylim(0,9)

violin1
```


```{r plot-errbar1}
errbar1 <- df %>%
  group_by(ft) %>%
  summarise(mean = mean(n_ind),
            sd = sd(n_ind),
            n = n(),
            SE = sd(n_ind) / sqrt(n())) %>%
  ggplot(aes(x = ft, y = mean)) +
  geom_point(shape = 15, size = 3) +
  geom_errorbar(aes(ymin = mean - SE, ymax = mean + SE), width = 0.2) +
  labs(title = "",
       x = "",
       y = "Mean ") +
  theme_classic() +
  annotate("text", x = 0.8, y = 0.9, label = "P = 0.982")

errbar1
```

```{r plot-inset1}
main.plot <- violin1

inset.plot <- errbar1

plot.with.inset <-
  ggdraw() +
  draw_plot(main.plot) +
  draw_plot(inset.plot, x = .65, y = .65, width = .4, height = .4)

plot.with.inset
```

Interesting that the higher outliers in the hemlock group make the mean higher for that group, while the medians of the two groups are the same.  Is median a good summary statistic when it is equal to zero?

Wilcoxon test for difference given the extreme skew.
```{r test-wilcox-n_ind-df}
wilcox.test(df$n_ind ~ df$ft)
```
Report: I found no significant difference in mean  per cover board between mixed deciduous and hemlock forests (Wilcoxon rank sum test, W = 1198, P = 0.982).

#### Site comparisons
```{r}
df %>%
  group_by(site) %>%
  summarise(mean = mean(n_ind),
            sd = sd(n_ind),
            n = n(),
            SE = sd(n_ind) / sqrt(n())) %>%
  ggplot(aes(x = site, y = mean)) +
  geom_point(shape = 15, size = 3) +
  geom_errorbar(aes(ymin = mean - SE, ymax = mean + SE), width = 0.2) +
  labs(title = "",
       x = "Site",
       y = "Mean abundance per cover board") +
  theme_bw()
```

```{r}
kw1 = kruskal.test(df$n_ind ~ df$site)
kw1
```
Significant, so we will use a Dunn's post hoc test.

```{r}
library(dunn.test)

dunn.test(df$n_ind, df$site)
```

Add report, not sure if this will go into presentation, but it is worth knowing differences between sites here.

### Modeling
#### Assess normality
```{r}
p1 <- ggplot(df, aes(x = ba)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Basal area (m2/ha)",
       y = "")
p2 <- ggplot(df, aes(x = st.ha)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Stems/ha",
       y = "")
p3 <- ggplot(df, aes(x = ts)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Percent hemlock",
       y = "Count")
p4 <- ggplot(df, aes(x = canopy.cover)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Percent canopy cover",
       y = "")
p5 <- ggplot(df, aes(x = vcwd.con)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Volume CWD (m3)",
       y = "")
p6 <- ggplot(df, aes(x = soil.ph)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(x = "Soil pH",
       y = "")

grid.arrange(p1, p2, p3, p4, p5, p6)
```


#### Assess relationships
```{r plot-relationships, message = FALSE, warning = FALSE}
p1 <- ggplot(df, aes(x = ba, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Basal area (m2/ha)",
       y = "")
p2 <- ggplot(df, aes(x = st.ha, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Stems/ha",
       y = "")
p3 <- ggplot(df, aes(x = ts, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Percent hemlock",
       y = "")
p4 <- ggplot(df, aes(x = canopy.cover, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Percent canopy cover",
       y = "Abundance per cover board")
p5 <- ggplot(df, aes(x = cwd.con, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Number CWD pieces contact",
       y = "")
p6 <- ggplot(df, aes(x = vcwd.con, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Volume CWD contact",
       y = "")
p7 <- ggplot(df, aes(x = avg.dc, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Average decay class",
       y = "")
p8 <- ggplot(df, aes(x = tot.snags, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Total snags",
       y = "")
p9 <- ggplot(df, aes(x = soil.ph, y = n_ind)) +
  geom_jitter(alpha = 0.25) +
  geom_smooth(method = "lm") +
  labs(x = "Soil pH",
       y = "")

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, top = grid::textGrob("Relationships between P. cinereus abundance and predictors"))
```

#### Hurdle model without interaction

```{r}
df1 <- na.omit(df)
hurdle_nb1 <- hurdle(n_ind ~ ft + ts + ba + st.ha + canopy.cover + cwd.con + vcwd.con + avg.dc + tot.snags + avg.snag.dbh + soil.ph, data = df1, dist = "negbin")
summary(hurdle_nb1)
```


```{r}
stepAIC(hurdle_nb1, direction = "both")
```
```{r}
hurdle_nb2 <- hurdle(n_ind ~ canopy.cover + avg.snag.dbh, data = df1, dist = "negbin")
summary(hurdle_nb2)
```
#### Hurdle model with forest type interaction
```{r}
hurdle_nb3 <- hurdle(n_ind ~ ft + ts + ba + st.ha + canopy.cover + cwd.con + vcwd.con + avg.dc + tot.snags + avg.snag.dbh + soil.ph + ft*ts + ts*ba + ts*soil.ph + ft*ba + ft*soil.ph, data = df1, dist = "negbin")
summary(hurdle_nb3)
```
```{r}
stepAIC(hurdle_nb3, direction = "both")
```

```{r}
plot_nb3_hurdle <- df %>%
  mutate(presence = case_when(n_ind >= 1 ~ 1, n_ind == 0 ~ 0)) %>%
  ggplot(aes(x = canopy.cover, y = presence)) +
  geom_jitter(height = 0.02, alpha = 0.25) +
  geom_smooth(method = "glm", method.args = list(family="binomial"), se = TRUE, fullrange = TRUE) +
  xlim(60,120) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(x = "",
       y = "")

plot_nb3_hurdle
```
```{r}
plot_nb3_count <- df %>%
  filter(n_ind > 0) %>%
  ggplot(aes(x = canopy.cover, y = n_ind)) +
  geom_jitter(alpha = 0.5) +
  stat_smooth(method = MASS::glm.nb, se = TRUE, fullrange = TRUE) +
  xlim(70,100) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(x = "Percent canopy cover",
       y = "Abundance per cover board")

plot_nb3_count
```

```{r}
main.plot <- plot_nb3_count

inset.plot <- plot_nb3_hurdle

hurdle_plot <-
  ggdraw() +
  draw_plot(main.plot) +
  draw_plot(inset.plot, x = 0.055, y = .58, width = .4, height = .4)

hurdle_plot
```

#### Hurdle model with multiple interaction
```{r}
hurdle_nb4 <- hurdle(n_ind ~ ft + ts + ba + st.ha + canopy.cover + cwd.con + vcwd.con + avg.dc + tot.snags + avg.snag.dbh + soil.ph + ft*ts + ts*ba + ts*soil.ph + ft*ba + ft*soil.ph + cwd.con * vcwd.con + st.ha * canopy.cover, data = df1, dist = "negbin")
summary(hurdle_nb4)
```
```{r}
stepAIC(hurdle_nb4, direction = "both")
```

```{r}
library(AICcmodavg)
models <- list(hurdle_nb1, hurdle_nb2, hurdle_nb3, hurdle_nb4)
mod.names <- c('Env model full', 'Final model', 'Env model ft interaction', 'Env model mult interactions')
aic_table <- aictab(cand.set = models, modnames = mod.names)
aic_table
```

```{r}
#install.packages("knitr")
library(knitr)
```
```{r}
kable(aic_table, col.names = c('Model', 'K', 'AIC', 'Delta AIC', 'Model Lik', 'AIC Wt', 'LL', 'Cum Wt'))
```

# Consider chi-squared test for absolute counts.


## P. cinereus Morphometrics

```{r}
wt_plot <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = wt, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "P. cinereus morphometrics by sex",
       x = "Weight (cg)",
       y = "",
       fill = "Sex") +
  theme(legend.position = c(0.96, 0.85)) +
  theme(legend.title = element_blank()) +
  scale_fill_viridis_d()

tl_plot <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = tl, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Total length (mm)",
       y = "Density") +
  theme(legend.position = "none") +
  scale_fill_viridis_d()

svl_plot <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = svl, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Snout-vent length (mm)",
       y = "") +
  theme(legend.position = "none")  +
  scale_fill_viridis_d()

grid.arrange(wt_plot, tl_plot, svl_plot)

#  ggpairs(aes(color = sex), columns = c("wt", "tl", "svl"))
```

Density distributions for Plethodon cinereus weight (in centigrams), total and snout-vent length (in millimeters).  Females appear to be somewhat heavier and longer than males.  Observations with missing values removed.  Skew in total length but not snout-vent length may be due to salamanders with missing/regenerating tails.

```{r boxplot-morpho-sex}
box1 <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = sex, y = wt, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "",
       y = "Weight (cg)") +
  scale_fill_manual(values = c("darkseagreen2", "lavender")) +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 162, label = "P < 0.001")

box2 <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = sex, y = tl, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "", 
       y = "Total length (mm)") +
  scale_fill_manual(values = c("wheat", "lavender")) +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 99, label = "P < 0.001")

box3 <- df_pc %>%
  drop_na(sex) %>%
  ggplot(aes(x = sex, y = svl, fill = sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "",
       y = "Snout-vent length (mm)") +
  scale_fill_manual(values = c("wheat", "lavender")) +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 56, label = "P < 0.001")

grid.arrange(box1, box2, box3, ncol = 3, top = textGrob("P. cinereus morphometrics by sex"), bottom = textGrob("Sex"))
```

Normally distributed, so we can use a T-test or simple linear model.

T-test first:
```{r}
t.test(wt ~ sex, data = df_pc)
```

I will also run T-tests for total length and snout-vent length as a function of sex.
```{r}
t.test(tl ~ sex, data = df_pc)
```
```{r}
t.test(svl ~ sex, data = df_pc)
```
Female P. cinereus are significantly heavier than males; total length and snout-vent length were also significantly longer in females than males.

#### By forest type
```{r boxplot-morpho-ft, warning = FALSE}
box1 <- df_pc %>%
  ggplot(aes(x = ft, y = wt, fill = ft)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "",
       y = "Weight (cg)") +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 162, label = "P = 0.069")

box2 <- df_pc %>%
  ggplot(aes(x = ft, y = tl, fill = ft)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "", 
       y = "Total length (mm)") +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 99, label = "P = 0.089")

box3 <- df_pc %>%
  ggplot(aes(x = ft, y = svl, fill = ft)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.1) +
  labs(x = "",
       y = "Snout-vent length (mm)") +
  theme_bw() +
  theme(legend.position = "none") +
  annotate("text", x = 2.1, y = 56, label = "P = 0.398")

grid.arrange(box1, box2, box3, ncol = 3, top = textGrob("P. cinereus morphometrics by forest type"), bottom = textGrob("Forest type"))
```

```{r}
t.test(wt ~ ft, data = df_pc)
wilcox.test(df_pc$wt ~ df_pc$ft)
```
No significant difference in mean weight between mixed deciduous and hemlock-dominated stands.

```{r}
t.test(tl ~ ft, data = df_pc)
wilcox.test(df_pc$tl ~ df_pc$ft)
```
No significant difference in mean total length between mixed deciduous and hemlock stands.

```{r}
t.test(svl ~ ft, data = df_pc)
wilcox.test(df_pc$svl ~ df_pc$ft)
```
No significant difference in mean snout-vent length between mixed deciduous and hemlock forests.

We can also model this with the continuous variable `ts` instead of categorical `ft`.
```{r}
lm_wt_ts <- linear_reg(engine = "lm") %>%
  fit(wt ~ ts, data = df_pc)

lm_wt_ts_tidy <- lm_wt_ts %>%
  tidy()

lm_wt_ts_tidy

glance(lm_wt_ts)
```
Not significant with very low R^2.

```{r}
lm_tl_ts <- linear_reg(engine = "lm") %>%
  fit(tl ~ ts, data = df_pc)

lm_tl_ts_tidy <- lm_tl_ts %>%
  tidy()

lm_tl_ts_tidy

glance(lm_tl_ts)
```
Significant (P = 0.02), but note very low R^2 (0.0089).

```{r}
lm_svl_ts <- linear_reg(engine = "lm") %>%
  fit(svl ~ ts, data = df_pc)

lm_svl_ts_tidy <- lm_svl_ts %>%
  tidy()

lm_svl_ts_tidy

glance(lm_svl_ts)
```
Also significant (P = 0.02), again with very low R^2 (0.0088).

#### Cluster Analysis
Compare total length, snout-vent length, and weight between sexes of P. cinereus
```{r}
#install.packages("cluster")
#install.packages("factoextra")
library(cluster)
library(factoextra)
```

```{r}
pc_morpho <- df_pc %>%
  dplyr::select(tl, svl, wt, sex) %>%
  drop_na(sex) %>%
  mutate(across(c(tl, svl, wt), scale)) %>% # Try with and without this line
  drop_na()

pc_means <- pc_morpho %>%
  dplyr::select(-sex)

k2 <- kmeans(pc_means, centers = 2, nstart = 25)
k3 <- kmeans(pc_means, centers = 3, nstart = 25)

## Calculate k4 and k5

k4 <- kmeans(pc_means, centers = 4, nstart = 25)
k5 <- kmeans(pc_means, centers = 5, nstart = 25)
```
Have to use `dplyr::select` because the `select` function is overwritten by another package.

```{r}
k2
```


```{r}
p1 <- fviz_cluster(k2, geom = "point", data = pc_means) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point",  data = pc_means) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point",  data = pc_means) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point",  data = pc_means) + ggtitle("k = 5")


grid.arrange(p1, p2,  p3, p4, nrow = 2)
p1
```

```{r}
fviz_nbclust(pc_means, kmeans, method = "wss")
```

```{r}
final <- k2

pc_means %>%
  mutate(sex = pc_morpho$sex,
  cluster = final$cluster) %>%
  group_by(cluster, sex) %>%
  summarize(count = n()) %>%
  ggplot() +
  geom_col(aes(x = sex, y = count, fill = as.factor(cluster)), color = "black") +
  scale_fill_manual(values = c("skyblue", "red4")) +
  coord_flip() +
  labs(x = "Sex",
       y = "Count",
       fill = "Cluster")
```

Try this by forest type / site.

```{r}
pc_morpho_ft <- df_pc %>%
  dplyr::select(tl, svl, wt, ft) %>%
  drop_na(ft) %>%
  mutate(across(c(tl, svl, wt), scale)) %>% # Try with and without this line
  drop_na()

pc_means_ft <- pc_morpho_ft %>%
  dplyr::select(-ft)

k2 <- kmeans(pc_means_ft, centers = 2, nstart = 25)
k3 <- kmeans(pc_means_ft, centers = 3, nstart = 25)

## Calculate k4 and k5

k4 <- kmeans(pc_means_ft, centers = 4, nstart = 25)
k5 <- kmeans(pc_means_ft, centers = 5, nstart = 25)
```

```{r}
p1 <- fviz_cluster(k2, geom = "point", data = pc_means_ft) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point",  data = pc_means_ft) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point",  data = pc_means_ft) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point",  data = pc_means_ft) + ggtitle("k = 5")


grid.arrange(p1, p2,  p3, p4, nrow = 2)
```

```{r}
fviz_nbclust(pc_means_ft, kmeans, method = "wss")
```
Seems like k = 2 probably is the better fit here.

```{r}
final <- k2

pc_means_ft %>%
  mutate(ft = pc_morpho_ft$ft,
  cluster = final$cluster) %>%
  group_by(cluster, ft) %>%
  summarize(count = n()) %>%
  ggplot() +
  geom_col(aes(x = ft, y = count, fill = as.factor(cluster))) +
  coord_flip() +
  labs(fill = "Cluster")
```

## Biodiversity
```{r}
herps <- read_csv("/cloud/project/data/hf131-06-herps.csv")

herps <- herps %>%
  mutate(herp_species = species)

herps %>%
  count(herp_species)
```
We also have 323 observations of individual P. cinereus we need to add here as well as check to see if there are any instances of multiple individuals of other herps in one observation.  Add PLCI to species with case_when or if_else.

```{r}
herps %>%
  filter(str_detect(osn, ","))
```
No commas in the unique identifiers assigned to other herp species, so we can assume there are no cases of multiple individuals per observation.

There is no variable for forest type assigned in the `herps` dataframe, so I will assign it based on the the plot code.

```{r}
herps <- herps %>%
  mutate(ft = str_extract(plot, "(?<=.)TS|MD"))

herps
```

I need to duplicate these rows with `pca` > 1 by the conditional value of `pca` so that each row is one individual P. cinereus observation.

`uncount` seems to work here.

```{r}
pc_obs <- df %>%
  uncount(pca)

pc_obs$herp_species = "PLCI"

pc_obs
```
323 rows all labeled by species (PLCI = Plethodon cinereus).

```{r}
sp <- full_join(pc_obs, herps)

sp <- sp %>%
  dplyr::select(ft, herp_species)
sp
```

```{r}
sp %>%
  count(herp_species) %>%
  arrange(desc(n))
```

```{r}
sp %>%
  group_by(ft) %>%
  count(herp_species) %>%
  arrange(herp_species)
```
Looks like all amphibian species were detected more frequently in hemlock stands than mixed deciduous; only the two snake species were more frequently detected in mixed deciduous stands.  Since snakes are a separate taxonomic group and are only found in one habitat, they are adding noise to the analysis.  We will remove them for analyzing biodiversity.

```{r}
amph <- sp %>%
  filter(herp_species %in% c("AMMA", "PLCI", "NOVI", "EUBI", "RASY"))

amph
```


```{r}
md <- amph %>%
  filter(ft == "MD")

md

ts <- amph %>%
  filter(ft == "TS")

ts
```

```{r create-richness-list}
md_matrix <- matrix(table(md))
md_abund <- as.abucount(md_matrix)
glimpse(md_abund)

ts_matrix <- matrix(table(ts))
ts_abund <- as.abucount(ts_matrix)
glimpse(ts_abund)

richness_list <- list(md_abund, ts_abund)
```

```{r}
richness <- iNEXT(richness_list, q = 0, datatype="")
richness
```
```{r}
ggiNEXT(richness, type=1, facet.var = "None", color.var = "Assemblage") +
  scale_color_discrete(name = "None", labels = c("Mixed deciduous", "Hemlock")) +
  scale_shape_discrete(name = "None", labels = c("Mixed deciduous", "Hemlock")) +
  scale_fill_discrete(name = "None", labels = c("Mixed deciduous", "Hemlock")) +
  labs(caption = "Amphibian species only")
```

Not particularly meaningful due to small sample size and sampling design targeted to detect P. cinereus, but it does suggest that amphibian biodiversity may be higher in hemlock stands.
