---
title: "Data Tidying"
author: "Joshua A. Harkness"
date: "2024-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-data-files, message = FALSE}
plots <- read_csv("/cloud/project/data/hf131-01-plot.csv")
trees <- read_csv("/cloud/project/data/hf131-02-tree.csv")
cwd <- read_csv("/cloud/project/data/hf131-03-cwd.csv")
aco <- read_csv("/cloud/project/data/hf131-04-aco.csv")
pc <- read_csv("/cloud/project/data/hf131-05-pc-abund.csv")
herps <- read_csv("/cloud/project/data/hf131-06-herps.csv")
```

```{r}
n_distinct(pc$pc.no)
```
250 uniquely identified P. cinereus according to this data file.  Does not include zeros so not useful for presence/absence.

```{r}
aco_abund <- aco %>%
  summarize(n_ind = sum(pca))

sum(aco_abund$n_ind)
```
241 individuals according to this data file, which includes all non-observations as zeroes.

Erroneous count data that I used previously resulted from joining these data files with `plots`, which apparently duplicated some rows and gave higher counts.

### Joining data files
```{r}

```

