---
title: "Data Tidying"
author: "Joshua A. Harkness"
date: "2024-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-data-files, message = FALSE}
plots <- read_csv("/cloud/project/data/hf131-01-plot.csv")
trees <- read_csv("/cloud/project/data/hf131-02-tree.csv")
cwd <- read_csv("/cloud/project/data/hf131-03-cwd.csv")
aco <- read_csv("/cloud/project/data/hf131-04-aco.csv")
pc <- read_csv("/cloud/project/data/hf131-05-pc-abund.csv")
herps <- read_csv("/cloud/project/data/hf131-06-herps.csv")
```

#### More complex join.

```{r}
aco %>%
  distinct(pcn)

pc %>%
  distinct(pc.no)
```
Since the `aco` dataframe is structure such that each row is a sampling event at a cover board with counts of P. cinereus, the authors lumped multiple unique identifiers for individual P. cinereus into one column (e.g. "115; 116").  I want these to match the values in the `pc.no` variable in the `pc` dataframe, which treats each P. cinereus as an observation.  Not super important since all observations are still grouped by plot and I don't plan to make comparisons below that level currently.

```{r}
aco %>%
  uncount(pca)
```

```{r}
count()
```

