---
title: "Red-backed salamander abundance and distribution in north-central Massachusetts"
author: "Joshua A. Harkness"
date: "2024-02-02"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(stringr)
library(sf)
library(ggspatial)
library(readxl)
```

## 1. Introduction



## 2. Data
### Load Packages

### Load Datafiles from GitHub Repository
```{r load-data-files, message = FALSE}
plots <- read_csv("/cloud/project/data/hf131-01-plot.csv")
trees <- read_csv("/cloud/project/data/hf131-02-tree.csv")
cwd <- read_csv("/cloud/project/data/hf131-03-cwd.csv")
aco <- read_csv("/cloud/project/data/hf131-04-aco.csv")
pc <- read_csv("/cloud/project/data/hf131-05-pc-abund.csv")
herps <- read_csv("/cloud/project/data/hf131-06-herps.csv")
```

```{r glimpse-data-files}
glimpse(plots)
glimpse(cwd)
glimpse(aco)
glimpse(pc)
glimpse(herps)
```

### Convert degrees, minutes, seconds to decimal degrees

```{r}
#latitude
plots <- plots %>%
  mutate(lat = str_replace_all(lat, "\xb0", ":"))

plots <- plots %>%
  mutate(lat = str_remove_all(lat, "N"))

plots <- plots %>%
  mutate(lat = str_remove_all(lat, "'"))

plots <- plots %>%
  mutate(lat = str_squish(lat))

#longitude
plots <- plots %>%
  mutate(long = str_replace_all(long, "\xb0", ":"))

plots <- plots %>%
  mutate(long = str_remove_all(long, "W"))

plots <- plots %>%
  mutate(long = str_remove_all(long, "'"))

plots <- plots %>%
  mutate(long = str_squish(long))
```

```{r}
#extract values to create separate columns for degrees, decimal minutes, and hemisphere


plots <- plots %>%
  mutate(lat_deg = str_extract_all(plots$lat, "..(?=:)"),
         lat_dm = str_extract_all(plots$lat, "(?<=:)......"),
         long_deg = str_extract_all(plots$long, "..(?=:)"),
         long_dm = str_extract_all(plots$long, "(?<=:)......"))

plots$lat_deg <- as.numeric(plots$lat_deg)
plots$lat_dm <- as.numeric(plots$lat_dm)
plots$long_deg <- as.numeric(plots$long_deg)
plots$long_dm <- as.numeric(plots$long_dm)
```

```{r}
coords <- plots %>%
  select(plot, lat_deg, lat_dm, long_deg, long_dm) %>%
  mutate(lat_dd = lat_deg + lat_dm / 60,
         long_dd = long_deg + long_dm / 60) %>%
  select(plot, lat_dd, long_dd)

coords$long_dd <- coords$long_dd*(-1)

plots$lat_dd <- coords$lat_dd
plots$long_dd <- coords$long_dd
```

```{r}
coords %>%
  ggplot(aes(x = long_dd, y = lat_dd)) +
  geom_point()
```

Based on the coordinates plotted here, data was collected within about fifteen miles of Harvard Forest in Petersham, Massachusetts.  The central clusters of points appear to be from the HF properties in Petersham; the southeastern cluster is from Mount Wachusett, and the northern and western clusters on state land in the towns of Northfield, Warwick, and Royalston.  This provides a useful reference as to the spatial scale of this project.

```{r}
mass_towns <- st_read("/cloud/project/data/Community Boundaries (Towns) from Survey Points.shp")
```
```{r}
ggplot(mass_towns) +
  geom_sf() #+
#  geom_point(data = coords, (aes(x = long_dd, y = lat_dd)), color = "black", alpha = 0.5)
```

Not sure why I'm getting this mismatch in coordinates.  My points appear in the correct location when plotted with X/Y coordinates, but they're showing up about 8 degrees too far south and a little west when plotted with this shapefile of Massachusetts towns.  Maybe because the points don't have a coordinate reference system assigned?  I haven't had this issue before though.  I'll try loading these points into ArcGIS to see if I figure out the source of the issue.


### Join data files
```{r}
# joining all except herps for now
plots_aco <- left_join(plots, aco)
plots_aco_cwd <- left_join(plots_aco, cwd)
df <- left_join(plots_aco_cwd, pc)
```


### Exploratory Visualizations

```{r}
df %>%
  ggplot(aes(x = site, y = soil.ph, fill = ft)) +
  geom_boxplot()
```

```{r}
df %>%
  ggplot(aes(x = site, y = ba, fill = ft)) +
  geom_boxplot()
```

```{r}
df %>%
  ggplot(aes(y = vcwd, x = ba)) +
  geom_col() +
  theme_bw() +
  labs(title = "Coarse Woody Debris and Basal Area",
       x = "Basal area (m2/ha)",
       y = "CWD volume (m3)")
```
```{r}
df %>%
  ggplot(aes(x = airt, y = soilt, shape = ft, color = ft)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Plethodon cinereus abundance

```{r}
df %>%
  count()
```


```{r}
df %>%
  ggplot(aes(x = tl.x, y = wt)) +
  geom_point() +
  geom_smooth()
```
```{r}
df %>%
  drop_na(sex) %>%
  ggplot(aes(x = sex, y = wt)) +
  geom_boxplot()
```

### Other herp morphometrics
```{r}
herps %>%
  count(species, sort = TRUE)
```


```{r}
herps %>%
  ggplot(aes(x = species, y = tl, fill = species)) +
  geom_boxplot() +
  coord_flip()
```


```{r}
df %>%
  ggplot(aes(x = vcwd, y = wt)) +
  geom_point()
```


## 3. Ethics review

## 4. Data analysis plan



